Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.

### 2024-03-21 article.tex
Results

Another challenge is the way models are defined.
Many tools support the \ac{SBML} \cite{huckaSBMLL3V2} as an exchange format,
a \emph{de facto} standard for \acp{CRN} that defines species, parameters and reactions between species.
As writing \ac{SBML} directly is impractical,
Tellurium uses a \ac{DSL} called Antimony \cite{smithAntimonyModularModel2009} for defining models.
\Acp{DSL} allows to reuse the same code in different programming environments,
but are not recognized by default in \acp{IDE}
and, therefore, they cannot provide syntax highlighting, code completion, refactoring, and static analysis.
For Antimony, an extension providing this capabilities was developed for Visual Studio Code,
but its maintenance could be a demanding task for the systems biology community.
In the case of PySB,
using Python's dynamic nature,
it developers designed a \ac{DSL} within Python.
By default, it uses the global scope to create species and parameters,
without explicitly assigning them to Python variables or to the model,
but this approach is not fully compatible with \acp{IDE},
affecting the development experience.

@@

Poincare is a package to define and simulate dynamical systems. It
provides a \texttt{System} class, where one can define
\texttt{Constant}s, \texttt{Parameter}s, \texttt{Variable}s, and create
equations linking a variable's derivative with an expression
(Figure~\ref{fig-first-order}). It also allows to define higher-order
systems by assigning an initial condition to a \texttt{Derivative}
(Figure~\ref{fig-second-order}).

Utilizing classes for system definition offers several advantages:

@@

To simulate a system, we created a \texttt{Simulator} instance
(Figure~\ref{fig-sim}) that compiles the \ac{RHS} of a given system and interfaces
with \ac{ODE} solvers. By default, it creates a first-order
\ac{ODE} system using \texttt{numpy} as a backend. This can be
easily switched to other solvers. The Simulator wraps the output in a
\texttt{pandas.DataFrame}, which can be easily plotted with the standard
\texttt{plot} method.

Switching backends to \texttt{"numba"} can improve model runtime by a
factor up to x30 or more for big or long running models, however incurs
in a compile time penalty for the first run that must be taken into
account.

>>>>

Another challenge is the way models are defined.
Many tools support the \ac{SBML} \cite{huckaSBMLL3V2} as an exchange format,
a \emph{de facto} standard for \acp{CRN} that defines species, parameters and reactions between species.
As writing \ac{SBML} directly is impractical,
Tellurium uses a \ac{DSL} called Antimony \cite{smithAntimonyModularModel2009} for defining models.
\Acp{DSL} allows to reuse the same code in different programming environments,
but are not recognized by default in \acp{IDE}
and, therefore, they cannot provide syntax highlighting, code completion, refactoring, and static analysis.
For Antimony, an extension providing this capabilities was developed for Visual Studio Code,
but its maintenance could be a demanding task for the systems biology community.
In the case of PySB,
using Python's dynamic nature,
it developers designed a \ac{DSL} within Python.
To save keystrokes,
it uses the global scope to create species and parameters,
without explicitly assigning them to Python variables or to the model,
but this approach is not fully compatible with \acp{IDE},
affecting the development experience.

@@

Poincare is a package to define and simulate dynamical systems.
It provides a \texttt{System} class,
where one can define
\texttt{Constant}s; which can be numbers or refer to other constants;
\texttt{Parameter}s, which can be numbers or time-dependent expressions;
\texttt{Variable}s, which represent the state of the system and must be provided with an initial condition;
and create equations linking a variable's derivative with an expression (Figure~\ref{fig-first-order}).
It also allows to define higher-order systems
by assigning an initial condition to a \texttt{Derivative} (Figure~\ref{fig-second-order}).

Within the constraints imposed by Python's current typing and static analyzers,
we define models utilizing Python classes
such that we can benefit from \acp{IDE}' autocomplete and refactoring capabilities.
This design offer several advantages:

@@

To simulate a system,
we created a \texttt{Simulator} instance (Figure~\ref{fig-sim})
that translates the model into the \ac{RHS} equations
and interfaces with \ac{ODE} solvers.
By default,
it creates a first-order \ac{ODE} system using \texttt{numpy} as a backend,
and uses the LSODA solver from \texttt{scipy}.
This can be easily switched to other solvers.
The Simulator wraps the output in a \texttt{pandas.DataFrame},
which can be easily plotted with the standard \texttt{plot} method.

Switching backends to \texttt{numba} can improve model runtime by a factor up to x30 or more for big or long running models,
however incurs in a compile time penalty for the first run that must be taken into account.
